Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-10-17 analysis/paper/paper.Rmd
more submission prep

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and to reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g.@flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

>>>>

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and to reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
highlights: |

  * Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.

  * We show how Bayesian integration of prior information from a national analysis of demographic-disturbance relationships can reduce uncertainty about the viability of sparsely monitored populations.

  * Integrating prior information about demographic-disturbance relationships clarifies that no additional information is needed to project population viability where disturbance is high and there are no actions to support recovery.

  * At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years is likely better than intensive short-term monitoring for reducing uncertainty about population viability.

  * We introduce a method for acknowledging potential errors in recruitment surveys to help clarify where and when these these errors are likely to alter conclusions about population viability.

editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].
